Epistemic Semantics in Guarded String Models

### 2020-09-23 tex/sdraft.tex
add change of state

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2020}
\usepackage{times}
\usepackage{latexsym}
%\usepackage{caption}
\renewcommand{\UrlFont}{\ttfamily\small}

@@

A running example of an event-sequence model is {\em The Concealed
  Coin}.  Amy and Bob are seated at a table.  There is a coin on the
table under a cup, heads up.  The coin could be heads or tails, and
neither agent knows which it is.  This initial situation is possible
world $w_1$.  Two additional worlds $w_2$ and $w_3$ are defined by
sequencing events after the initial state, with events interpreted as
in \ex{coin1}. The truth values for English sentences shown in
\ex{coin3} are observed.

\begin{example}  \label{coin1}
\hspace{-4mm} \begin{tabular}[t]{ll}
  $a_1$  & Amy peeks at heads, by tipping the
  \\ & cup. Bob sees she's peeking, but not\\
  & what she sees. \\
  $b_1$  & Bob peeks at heads. \\
  $a_0$  & Amy peeks at tails. \\
  $b_0$  & Bob peeks at tails. \\     
\end{tabular}
\end{example}

\begin{example} \label{coin2}
   $w_2 = w_1 \hspace{2pt} a_1 $ \hspace{2mm}
   $w_3 = w_1 \hspace{2pt} a_1 \hspace{2pt} b_1 $ 
\end{example}

\begin{example} \label{coin3}
\hspace{-9mm} \begin{tabular}[t]{cccl}
  $w_1$ & $w_2$  & $w_3$ & Sentence\\
  false & true & true& Amy knows that it's \\ &&& heads. \\
  false & false & true& Bob knows thats it's \\&&&heads. \\
  false & false & true& Bob knows Amy \\ &&& knows it's heads. \\
  false & true & true& Bob knows Amy \\ &&& knows  whether it's \\ &&& heads or tails. \\
\end{tabular}
\end{example}

\newcommand{\ev}[2]{#1 \!\! : \!\! #2}
\newcommand{\evr}[2]{#2 \!\! : \!\! #1}

The events come with pre-conditions.
Amy can peek at heads only if the coin is heads up, so $a_1$ has the precondition of
the coin being heads up.  Let $h$ be the Boolean proposition that the
coin is heads up.
Then preconditions
can be described by Boolean formulas, with $h$ being the precondition
of $a_1$
$u$. Events come as well with a relation between
prior and following state, for instance with $u$ incrementing the
floor.  This is expressed using an operator ``$:$'' (read ``and
next'') that pairs Boolean formulas. The formula $h:h$
describes $a_1$ (Amy looking at heads) as happening only in an $h$
state, and as not changing the state.  Symmetrically, $a_0$ (Amy
looking at tails) can happen only in a not-$h$ state, and does not
change the state ($\ev{\bar{h}}{\bar{h}}$).

@@

Suppose that in the coin example, we have an additional primitive
stative proposition (or atomic test) $t$, interpreted as tails.  A
sequence such as $\bar{h}{t}$ can be viewed as a valuation of
primitive propositions, which is used to describe world state. The
primitives are listed in fixed order, and left unmarked (indicating
true) or marked with the overbar (indicating false).  Since a coin is
heads or tails but not both, we want to allow the valuations
$h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and $\bar{h}\bar{t}$.
This is enforced by a {\em state formula}, which is a Boolean formula,
in this case the one given on the second line of \ex{effect}.  Where
{\sf{B}} is a set of atomic tests and $\phi$ is a state constraint
over {\sf{B}}, ${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\phi}$ is the set of
valuations of {\sf{B}} that make formula $\phi$ true. Valuations are
called atoms, because they correspond to the atoms of a Boolean
algebra of tests \citep{kozen2001automata}.

@@

Epik is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epik program that describes a possible worlds model for two agents
with information about one coin, and events of the agents
semi-privately looking at the coin. The line beginning with {\tt
  state} enumerates ${\sf{B}}$. The line beginning with {\tt
  restrict} gives the state formula.  The lines beginning with {\tt
  event} declare events and their effect formulas.  Finally the lines
beginning with {\tt agent} define {\em event alternative} relations
for agents.  Each clause with an arrow has a single event symbol on
the left, and a disjunction of alternative events on the right of the
arrow. The interpretation of Amy's alternatives for $b_1$ (Bill peeks
at heads), is that when $b_1$ happens, for Amy either $b_1$ or $b_0$
(Bill peeks at tails) could be happening.

\begin{figure}
  \begin{tcolorbox}
    \footnotesize
    \begin{minipage}{0.45\textwidth}
      \begin{center}    
        {\tt
          \begin{tabular}{l}
            state h t \\
            restrict h!t + t!h \\
            event a1 h:h \\
            event a0 t:t \\
            event b1 h:h \\
            event b0 t:t \\
      \end{tabular}}
      \end{center}
    \end{minipage} \hfill 
    \begin{minipage}{0.45\textwidth}
      \begin{center}
        {\tt
          \begin{tabular}{l}
            agent aly \\
            \hspace{3mm} a1 -> a1 \\
            \hspace{3mm} a0 -> a0 \\
            \hspace{3mm} b1 -> b1 + b0\\
            \hspace{3mm} b0 -> b1 + b0\\
            agent bob \\
            \hspace{3mm} b1 -> b1 \\
            \hspace{3mm} b0 -> b0 \\
            \hspace{3mm} a1 -> a1 + a0\\
            \hspace{3mm} a0 -> a1 + a0\\
        \end{tabular} }
      \end{center}
    \end{minipage}
  \end{tcolorbox}

\caption{Epik program describing a possible-worlds event sequence model for two agents
  with information about one coin, and events of the agents semi-privately looking
  at the coin.}
\label{fig:epik-example}
\label{epikfigure}
\end{figure}

@@

Guarded strings over a finite alphabet ${\sf{P}}$
are like ordinary strings, but with atoms over a set ${\sf{B}}$  
alternating with the symbols
from ${\sf{P}}$.  In the algebra described by Figure \ref{epikfigure}, 
${\sf{P}}$ is the set of events $\{a_1,a_0,b_1,b_0\}$, and
${\sf{B}}$ is $\{h,t\}$.

@@

\hspace{-2mm}\(H \; \diamond \; T \,a_1\, T = \textit{undefined} \)
\end{example}

@@

With modalities and propositional negation added, the signature of $n$-agent
epistemic KAT is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1
  ... \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet^*$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet^* h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet^* h$ is the set
of worlds where the coin is heads.\footnote{A mapping from English sentences to logical
  terms in epistemic KAT is presented in Section 6}.

@@

\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\bullet^*\neg t)$\\
 Bob knows that Amy knows whether \\ \hspace{3mm} it's tails.  \\
\end{tabular}
\end{example}

@@

\begin{examples} \label{three}
\itema Amy knows that Bob knows that Amy knows whether it is heads and knows that Bob doesnt know that Amy knows that it is tails.
%% \itemb {\tt (And (Not (Diamond amy (Not (Not (Diamond bob (Not (Or (Not (Diamond amy (Not heads))) (Not (Diamond amy heads))))))))) (Not (Diamond amy (Not (Not (Not (Diamond bob (Not (Not (Diamond amy (Not tails))))))))))) }
%% \itemc
%% \(\neg \Diamond_{\text{Amy}} \neg \neg \Diamond_{\text{Bob}} \neg \neg (\Diamond_{\text{Amy}}\bar h + \neg \Diamond_{\text{amy}} h) \cdot
%% \neg \Diamond_{\text{Amy}} \neg \neg \neg \Diamond_{\text{Bob}} \neg \neg \Diamond_{\text{Amy}} \bar t\)
\itemb \(\displaystyle \Box_{\text{Amy}}\Box_{\text{Bob}}(\Box_{\text{Amy}} \bar h + \Box_{\text{Amy}}\bar t)\) \hfill \mbox{} \(\cdot~~\Box_{\text{Amy}}\neg\Box_{\text{Bob}}\Box_{\text{Amy}}\bar t\)

@@

\bibliography{anthology,acl2020}
\bibliographystyle{chicago}%{acl_natbib}

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{acl2020}
\usepackage{times}
\usepackage{latexsym}
\usepackage{enumitem}
%\usepackage{caption}
\renewcommand{\UrlFont}{\ttfamily\small}

@@

A running example of an event-sequence model is {\em The Concealed
  Coin}.  Amy and Bob are seated at a table.  There is a coin on the
table under a cup, heads up.  The coin could be heads-up $H$ or
tails-up $T$, and neither agent knows which it is.  This initial
situation is possible world $w_1$.  Two additional worlds $w_2$ and
$w_3$ are defined by sequencing events after the initial state, with
events interpreted as in \ex{coin1}. The truth values for English
sentences shown in \ex{coin3} are observed, where $0$ stands for
falsity and $1$ for truth.

\newcommand{\apH}{\ensuremath{a_1}}
\newcommand{\apT}{\ensuremath{a_0}}
\newcommand{\bpH}{\ensuremath{b_1}}
\newcommand{\bpT}{\ensuremath{b_0}}

\newcommand{\atTH}{\ensuremath{a_{01}}}
\newcommand{\atHT}{\ensuremath{a_{10}}}
\newcommand{\btTH}{\ensuremath{b_{01}}}
\newcommand{\btHT}{\ensuremath{b_{10}}}

\begin{example}  \label{coin1} \begin{enumerate}[align=left,leftmargin=0.75em]
  \item[\apH] Amy peeks at $H$, by tipping the cup. Bob sees she's peeking, but not what she sees.
  \item[\bpH] Bob peeks at $H$.
  \item[\apT] Amy peeks at $T$.
  \item[\bpT] Bob peeks at $T$. 
  \item[\atTH] Amy secretly turns the coin from $T$ to $H$.
    She knows she turned the coin over, but not which side was face up. Bob thinks nothing happened.
  \item[\atHT] Amy secretly turns the coin from $H$ to $T$.
  \item[\atTH] Bob secretly turns the coin from $T$ to $H$.
  \item[\btHT] Bob secretly turns the coin from $H$ to $T$.
\end{enumerate}
\end{example}

\begin{example} \label{coin2}
  \(w_2 = w_1\apH \hfill
  w_3 = w_2\bpH \hfill
  w_4 = w_3\atHT\btTH\bpH\)
\end{example}

\begin{example} \label{coin3}
\hspace{-7mm} \begin{tabular}[t]{ccccl}
  $w_1$ & $w_2$  & $w_3$ & $w_4$ & Sentence\\
  $0$ & $1$ & $1$& $0$ & Amy knows it's heads. \\
  $0$ & $0$ & $1$& $1$ & Bob knows it's heads. \\
  $0$ & $0$ & $1$& $0$ & Bob knows Amy \\ &&&& knows it's heads. \\
  $0$ & $1$ & $1$& $0$ & Bob knows Amy \\ &&&& knows  whether it's \\ &&&& heads or tails. \\
\end{tabular}
\end{example}

\ehc{I'm not sure if these sentences are the most interesting anymore!}

\newcommand{\ev}[2]{#1 \mathbin{:} #2}
\newcommand{\evr}[2]{\ev{#2}{#1}}

The events come with pre-conditions and post-conditions.  Amy can turn
the coin from heads to tails only if the coin is heads-up, so $\atHT$
has the pre-condition of the coin being heads up. Once she turns the
coin over, tails must be face-up, so $a_{10}$ has the post-condition
of the coin being tails-up. Let $h$ be the Boolean proposition that
the coin is heads up and $t$ be the Boolean proposition that the coin
is tails-up.  Then pre- and post-conditions can be described by
Boolean formulas, with $h$ being the pre-condition of $\atHT$ and
$a_{01}$ being the post-condition. This is expressed using an operator
``$:$'' (read ``and next'') that pairs Boolean formulas. The formula
$\ev h t$ describes $\atHT$ (Amy turning the coin from heads
to tails) as happening only in an $h$ state, and concluding in a
not-$h$ state.  Events don't have to change state: the event $\apT$
(Amy peeking at tails) can happen only in a $t$ state, and does
not change the state ($\ev{t}{t}$).

@@

However a coin cannot be showing both heads and tails! Currently the
precondition $h$ of $\atHT$ only says that heads must be showing, and
says nothing about the fact that tails must be face-down, indicated by
the formula $\bar t$. We will further restrict the feasible conditions
for our actions by restricting the space of valuations for our formulae.

A sequence such as $\bar{h}{t}$ can be viewed both as a formula and as
a valuation of primitive propositions, which we use to describe world
state. The primitives are listed in fixed order, and left unmarked
(indicating true) or marked with the overbar (indicating false).
Since a coin is heads or tails but not both, we want to allow the
valuations $h\bar{t}$ and $\bar{h}t$, and disallow $ht$ and
$\bar{h}\bar{t}$.  This is enforced by a {\em state formula}, which is
a Boolean formula, in this case the one given on the second line of
\ex{effect}.  Where {\sf{B}} is a set of atomic tests and $\phi$ is a
state constraint over {\sf{B}},
${\cal{A}}_{\mbox{\sf\tiny{B}}}^{\phi}$ is the set of valuations of
{\sf{B}} that make formula $\phi$ true. Valuations are called atoms,
because they correspond to the atoms of a Boolean algebra of tests
\citep{kozen2001automata}.

@@

Epik is a specification language for possible worlds models that
includes declarations of events and states, state formulas, effect
formulas, and additional information.  Figure \ref{epikfigure} shows
an Epik program that describes a possible worlds model for two agents
with information about one coin, and events of the agents
semi-privately looking at the coin. The line beginning with {\tt
  state} enumerates ${\sf{B}}$. The line beginning with {\tt restrict}
gives the state formula.  The lines beginning with {\tt event} declare
events and their effect formulas.  Finally the lines beginning with
{\tt agent} define {\em event alternative} relations for agents.  Each
clause with an arrow has a single event symbol on the left, and a
disjunction of alternative events on the right of the arrow. The
interpretation of Amy's alternatives for $b_1$ (Bob peeks at heads),
is that when $b_1$ happens, for Amy either $b_1$ or $b_0$ (Bob peeks
at tails) could be happening. Her alternatives for $a_{01}$ and
$a_{10}$ (she turns the coin over) are $a_{10}$ and $a_{01}$,
indicating that she doesn't know, \emph{a priori}, whether she's
turning the coin from $H$ to $T$ or from $T$ to $H$. Similarly, Bob
secretly turns the coin over, in $b_{10}$ or $b_{01}$, she doesn't
know anything has happened, so her alternatives are the no-operation
events $o_1$ and $o_0$ for heads-worlds and tails-worlds
respectively. Bob's alternatives are the same, \emph{mutatis mutandi}.

\begin{figure}
  \begin{tcolorbox}
    \footnotesize
    \begin{minipage}{0.48\textwidth}
      \begin{center}    
        {\tt
          \begin{tabular}{l}
            state h t \\
            restrict  h!t \\
            \phantom{restrict }+ t!h \\
            event o1  h:h \\
            event o0  t:t \\
            event a1  h:h \\
            event a0  t:t \\
            event b1  h:h \\
            event b0  t:t \\
            event a10 h:t \\
            event a01 t:h \\
            event b10 h:t \\
            event b01 t:h \\
      \end{tabular}}
      \end{center}
    \end{minipage} \hfill 
    \begin{minipage}{0.51\textwidth}
      \begin{center}
        {\tt
          \begin{tabular}{l}
            agent aly \\
            \hspace{3mm} o1  -> o1 \\
            \hspace{3mm} o0  -> o0 \\
            \hspace{3mm} a1  -> a1 \\
            \hspace{3mm} a0  -> a0 \\
            \hspace{3mm} b1  -> b1 + b0\\
            \hspace{3mm} b0  -> b1 + b0\\
            \hspace{3mm} a10 -> a10 + a01 \\
            \hspace{3mm} a01 -> a10 + a01 \\
            \hspace{3mm} b10 -> o0 + o1 \\
            \hspace{3mm} b01 -> o0 + o1 \\
            agent bob \\
            \hspace{3mm} <\textrm{\textit{sim.} swap }a\textrm{ and }b>
            %% agent bob \\
            %% \hspace{3mm} b1 -> b1 \\
            %% \hspace{3mm} b0 -> b0 \\
            %% \hspace{3mm} a1 -> a1 + a0\\
            %% \hspace{3mm} a0 -> a1 + a0\\
        \end{tabular} }
      \end{center}
    \end{minipage}
  \end{tcolorbox}

\caption{Epik program describing a possible-worlds event sequence
  model for two agents with information about one coin, and events of
  the agents semi-privately looking at the coin. }
\label{fig:epik-example}
\label{epikfigure}
\end{figure}

@@

Guarded strings over a finite alphabet ${\sf{P}}$
are like ordinary strings, but with atoms over a set ${\sf{B}}$  
alternating with the symbols
from ${\sf{P}}$.  In the algebra described by Figure \ref{epikfigure}, 
${\sf{P}}$ is the set of events $\{a_1,a_0,b_1,b_0,a_{10},a_{01},b_{10},b_{01}\}$, and
${\sf{B}}$ is $\{h,t\}$.

@@

\hspace{-2mm}\(T\,b_{01}\,H \; \diamond \; T \,a_1\, T = \textit{undefined} \)
\end{example}

@@

With modalities and propositional negation added, the signature of $n$-agent
epistemic KAT is $\tuple{K,+,\cdot,*,\bar{},0,1,\neg,\Diamond_1
  ... \Diamond_n}$. Figure \ref{termfigure} defines the syntax of the language.
Juxtaposition is used for product.  Terms in this language
are used to represent the proposositional semantic values of English sentences.
\ex{interpretation1} gives some examples.  To explain the first one $\bullet$ as
defined in Figure \ref{fig:epik-grammar} is the disjunction of the primitive events.
Since a world is a well-formed sequence of events, $\bullet^*$ is the set of worlds.
Multiplying by the state symbol $h$ in the term $\bullet^* h$ has the effect of
conjoining $h$ with the atom at the end of the world.  So $\bullet^* h$ is the set
of worlds where the coin ends heads-up\footnote{A mapping from English sentences to logical
  terms in epistemic KAT is presented in Section 6}.

@@

\begin{tabular}[t]{l}
  $\Box_b(\Box_a\bullet^*t + \Box_a\neg\bullet^*t)$\\
 Bob knows that Amy knows whether \\ \hspace{3mm} it's tails.  \\
\end{tabular}
\end{example}

@@

\begin{examples} \label{three}
\itema Amy knows that Bob knows that Amy knows whether it is heads and knows that Bob doesnt know that Amy knows that it is tails.
%% \itemb {\tt (And (Not (Diamond amy (Not (Not (Diamond bob (Not (Or (Not (Diamond amy (Not heads))) (Not (Diamond amy heads))))))))) (Not (Diamond amy (Not (Not (Not (Diamond bob (Not (Not (Diamond amy (Not tails))))))))))) }
%% \itemc
%% \(\neg \Diamond_{\text{Amy}} \neg \neg \Diamond_{\text{Bob}} \neg \neg (\Diamond_{\text{Amy}}\bar h + \neg \Diamond_{\text{amy}} h) \cdot
%% \neg \Diamond_{\text{Amy}} \neg \neg \neg \Diamond_{\text{Bob}} \neg \neg \Diamond_{\text{Amy}} \bar t\)
\itemb \(\displaystyle \Box_{\text{Amy}}\Box_{\text{Bob}}(\Box_{\text{Amy}} \neg \bullet^* h + \Box_{\text{Amy}}\neg \bullet^* t)\) \hfill \mbox{} \(~~\cdot~~\Box_{\text{Amy}}\neg\Box_{\text{Bob}}\Box_{\text{Amy}}\neg \bullet^* t\)

@@

\bibliography{anthology,acl2020}
\bibliographystyle{acl_natbib}
